---
title: "Rush Hour Explorations"
subtitle: "Assignment for PVG0038 Reproducibility in Research with a focus on data analysis using the program R"
author: "Alexander Koc"
date: '2020-11-05'
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

In this short report I investigate the value of adding two new features to the Kaggle bikeshare data set: `hour` denoting 
hour of day and boolean variable `rush_hour` to denote if the current hour is during rush hour or not. The two variables
will be evaluated together with the remaining predictors by fitting a random model forest and extracting variable 
importance metrics.

## Materials and Methods 

## Results

### Data preparation

Set random seed for reproducibility.

```{r}
set.seed(777)
```

To install all packages required for analysis, please run the following command in R `source("install_dependencies.R")`

```{r libs_and_data, message=FALSE}
library(tidyverse)
library(corrplot)
library(caret)
library(lubridate)
library(here)
library(parallel)
library(ranger)
library(forcats)

# Set number of cores for parallel computation
nbr_cores <- parallel::detectCores()

# Create variable specification for dataset:
col_spec <- cols(
  .default = col_double(),
  datetime = col_datetime(format = ""),
  season = col_integer(),
  holiday = col_logical(),
  workingday = col_logical(),
  weather = col_factor(),
  count = col_integer()
)

bikeshare <- read_csv(here::here("data/train.csv"), col_types = col_spec)
```

We drop the `casual` and `registered` columns, as they add up to the `count` column anyway -- making it a linear 
combination of both.

```{r}
bikeshare <- bikeshare %>% 
  select(-registered, -casual)
```

Expand season names to text labels -- we have the computer memory and screen real estate to spare:

```{r}
bikeshare <- bikeshare %>% 
  mutate(season = map_chr(season, function(season_code) {
    
    seasons <- c("spring" = 1, "summer" = 2,  "fall" = 3, winter = "4")
    translated_name <- names(seasons[season_code])
    
    # (I guess map_chr would catch this anyway?)
    if(is.na(translated_name)) {
      err <- paste("Invalid season code:", season_code)
      stop(err)
    }
    
    return(translated_name)
    
  }))

# Convert back to factor:
bikeshare$season <- as.factor(bikeshare$season) 
```

Expand some column names to spell out what they actually are and rename others to make data set look nicer. 

```{r}
bikeshare <-  bikeshare %>% 
  rename(
    date_time = datetime,
    working_day = workingday,
    temp_celcius = temp,
    feels_like_temp_celcius = atemp,
    relative_humidity = humidity,
    wind_speed = windspeed,
    total_nbr_rentals = count
  )
```

Finally, the hour at which bikes are rented _feels_ like it should have some predictive power. Split the date time
into a time column:

```{r}
bikeshare <- bikeshare %>% 
  mutate(hour = hour(date_time)) %>%
  select(date_time, hour, everything())
```

Probably good idea to create another column that specified which day of the week it is. A weekend starts after working
hours on a friday, and I hypothesize there will be more bike sharing activity on a friday evening -> midnight compared
to other weekdays. Out of the scope for this assignment though.

Final data set:

```{r}
glimpse(bikeshare)
```

### Data exploration

#### Data summary

```{r}
summary(bikeshare)
```
Seems like there is only one row with weather code "4". Remove it:

```{r}
bikeshare <- bikeshare %>%
  filter(weather != 4)

# Drop unused factor level to avoid annoying messages later:
bikeshare$weather <- forcats::fct_drop(bikeshare$weather)
```

Check for correlation between counts and numeric variables
 
```{r}
cor_mat <- bikeshare %>% 
  select(where(is.numeric), -hour) %>% 
  cor(method = "spearman") %>% 
  corrplot()
```
The two temperature variables are practically the same:

```{r}
bikeshare %>% 
  ggplot(aes(y = feels_like_temp_celcius, x = temp_celcius)) + 
  geom_point() 
```

Uhh, except that really weird horizontal line. What's going on? Let's look at that interval.

```{r}
bikeshare %>% 
  filter(temp_celcius > 25 & temp_celcius < 36) %>% 
  filter(feels_like_temp_celcius < 25) %>% 
  select(date_time, temp_celcius, feels_like_temp_celcius)
  
```

Seems like a wrong value was propagated for the entire day. Here I would try to delete that the wrong temperature values 
and impute new values. No time for that though -- this time we just delete:

```{r}
bikeshare <- bikeshare %>% 
  filter(!(temp_celcius > 25 & temp_celcius < 36 & feels_like_temp_celcius > 10 & feels_like_temp_celcius < 15))

bikeshare %>% 
  ggplot(aes(y = feels_like_temp_celcius, x = temp_celcius)) + 
  geom_point() 
```
#### Missingness

Check for missing data:
 
```{r}
# Returns true is there are _any_ missing values in the data frame
bikeshare %>% 
  is.na %>% 
  any()
```
No missing data.


### Adding new features

#### `hour`

Visualize total number of rentals by daily hour -- the new variable introduced earlier in code. 

```{r}
bikeshare %>% 
  ggplot(aes(x = hour, y = total_nbr_rentals)) +
  geom_jitter(alpha = 0.1)
```

Clear peak of rides around rush hours. Investigate to see if there is an interaction
with the workday variable.

```{r}
bikeshare %>% 
  ggplot(aes(x = hour, y = total_nbr_rentals, color = working_day)) +
  geom_jitter(alpha = 0.1)
```

Split up the plot above into two plots and the pattern becomes more clear. Rush hour peaks of course exist on working days.
Weekends seem to have more bike share activity in the hours on and after midnight.

```{r}
bikeshare %>% 
  ggplot(aes(x = hour, y = total_nbr_rentals)) +
  geom_jitter(alpha = 0.1) +
  facet_wrap(~working_day)
```

#### `rush_hour`

Create new boolean variable `rush_hour` by intersecting all hours between 05-10 and 15-19 when `working_day == TRUE`:

```{r}
bikeshare <- bikeshare %>% 
  mutate(rush_hour = map2_lgl(hour, working_day, function(hour, working) {
    
    # If not working day then no rush hour
    if(working == FALSE) {
      return(FALSE)
    }
    
    if((5 < hour & hour < 10) | (15 < hour & hour < 19)) {
      return(TRUE)
    }
    
    return(FALSE)
    
  }))
```

How does the new feature describe the data?

```{r}
bikeshare %>% 
  ggplot(aes(fill = rush_hour, x = total_nbr_rentals)) + 
  geom_histogram()

```
Less counts distributed around 0 for rush hour hours.

#### Interactions of new features with other features

Continue looking at rides per hour, but this time split by seasons and weather

```{r}
bikeshare %>% 
  ggplot(aes(x = hour, y = total_nbr_rentals)) +
  geom_jitter(alpha = 0.1) + 
  geom_smooth(method = "loess") + 
  facet_grid(rows = vars(working_day), cols = vars(season))
```
Again, what is it about spring? Not much visual difference between other seasons. Even seems like spring has effect on weekday
bike shares.

Next up, split by weather.

```{r}
bikeshare %>% 
  ggplot(aes(x = hour, y = total_nbr_rentals)) +
  geom_jitter(alpha = 0.1) + 
  geom_smooth(method = "loess") + 
  facet_wrap(~working_day + weather) 
```
Unbalanced data with worsening weather, but _if_ plots are to be believed then less rides in bad weather.

Could look more into how weather types interact with seasons.

Look at temperature:

```{r}
bikeshare %>% 
  ggplot(aes(x = temp_celcius, y = total_nbr_rentals)) + 
  geom_point(alpha = 0.1) +
  geom_smooth() 
```

```{r}
bikeshare %>% 
  ggplot(aes(x = temp_celcius, y = total_nbr_rentals)) + 
  geom_point(alpha = 0.1) +
  geom_smooth() +
  facet_wrap(~hour)
```
The effect of temperature is specific to which hour it is, sort of. Guessing there is _probably_ an interaction with 
season and weather, not to mention humidity here. Outside the scope of this course though.

How about wind speed?


```{r}
bikeshare %>% 
  ggplot(aes(x = wind_speed, y = total_nbr_rentals)) + 
  geom_point(alpha = 0.1) +
  geom_smooth() +
  facet_wrap(~hour)
```
Nothing as clear as the temperature effect by hour.

### Model

We will fit a random forest model using caret. Best model will be selected on basis of 5-fold cross validation.
Variable importance will be assessed by extracting variable importance metrics from final model and plotting.

```{r}
train_control <- trainControl(method = "repeatedcv", number = 5, repeats = 0, verboseIter = FALSE, allowParallel = TRUE)
```

Remove the `date_time` column and fit model using all predictor variables. Use `ranger` for parallel random forest 
model fit. Predictors are centered and scaled.

```{r caret_model_fit, cache=TRUE}
# This might take a few minutes to run!
fit_mod <- bikeshare %>% 
  select(-date_time) %>% 
  train(total_nbr_rentals ~ ., data = ., method = "ranger", trControl = train_control, preProcess = c("center", "scale"), num.threads = nbr_cores, importance = "impurity")

fit_mod
```
Extract and plot the variable importance metrics for predictors:

```{r}
variable_importance <- varImp(fit_mod)
plot(variable_importance)
```
The two added features `hour` and `rush_hour` are at the top of the variable importance plot.

## Literature

## Session

Packages and R version used in my analysis.

```{r}
devtools::session_info()
```

